\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-06-02 eg6-letters/NIM/EG6_RTPC_NIM.tex
Edits from RD to NIM

The CLAS RTPC is a 250~mm long cylinder of 158~mm diameter, leaving just enough 
room to fit pre-amplifiers between the RTPC outer shell and the solenoid. The 
electromagnetic field is directed perpendicularly to the beam direction, 
such that drifting electrons are pushed away from the beam line. These electrons 
are amplified by three layers of curved GEMs and detected by the readout system 
on the external shell of the detector as illustrated in the figure~\ref{fig:RTPC_1_4}. The CLAS 
RTPC is segmented into two halves with independent GEM amplification systems that 
cover about 80\% of the azimuthal angle.

We detail here the different regions shown in figure~\ref{fig:RTPC_1_4} starting 
from the beam line and towards larger radius:\\
\begin{itemize}
   \item First, the 6 atm helium-4 target extends along the beam line axis, it 
      is 284~mm long with a 3~mm radius wall made of a 27-$\mu$m-thick Kapton.
   \item The first gas gap covers a radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o ller electrons scattered by the beam. This 
      region is surrounded by a 4 $\mu$m thick window made of aluminized Mylar 
      is electronically grounded.
   \item The second gap region extends between 20~mm and 30~mm and is filled with a 
      gas mixture of 80$\%$ neon and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar, which 
      serves as the cathode and is set at 4260~V.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first Gas Electron Multiplier(GEM), 60 mm away 
      from the beam axis. The average electric field in this region is about 550~V/cm 
      and perpendicular to the beam.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied to each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Preamplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

Due to the non perfect experimental conditions, in particular variations in the 
gas mixture\footnote{It was noted during the experiment that leaks appeared
and the gas flow had to be increased in few occasions to compensate for this.}, 
the drift speed changes during the three months long experimental run.  
Figure \ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for 
individual runs (approximately 2 hours long). We observe a significant 
variation of the drift speed over time and accounted for it in the drift 
speed used for the track reconstruction.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. Software exists to calculate the drift paths, in particular
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, gas composition is far from stable in the 
chamber. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be around the millimeter precision. 
These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These are taken with a lower energy 
electron beam (1.20 and 1.27 GeV) to enhance the cross section of the elastic 
process ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
electron kinematics allows to calculate the Helium nucleus kinematic. It is
by comparing this calculated kinematic to the measured one that we fine tune
the drift paths independently of our knowledge of the exact conditions in the 
chamber.

The drift paths are ajusted using a set of identified elastic events 
from our lower beam energy run. Based on the kinematic of the electrons in
these events, we generate the helium nucleus in a GEANT4 
simulation~\cite{GEANT4} of our RTPC. Then we compare the calculated GEANT4 trajectory of 
the Helium nuclei to the hits measured in the chamber.

Because of the 
magnetic field, the drift paths are not linear in the RTPC. So to perform the extraction, 
we make a first approximation with a linear dependence between the radius of 
emission and the time of detection, and then refine our result. As it happens, 
the curvature is minimal and this process converge already on the second 
iteration.

At the end of the extraction procedure, where we work in a polar coordinate 
system ($R$,$\phi$), the azimuthal difference between the detection pad and 
the ionization point ($\Delta\phi$) is extracted as a function of time. 
In figure~\ref{fig:DELTA_PHI_TDC}, we show the resulting data points for one 
bin, where the drift paths is easily identified and eventually fitted for 
implementation in our reconstruction codes.

@@

To verify the stability of the drift paths, this procedure was carried out 
using both the 1.204 GeV data from the beginning of the run period and the 
1.269 GeV data from the end of the run period (shown in blue on 
figure~\ref{fig:Drift_run_number_1}). Strangely, we found very similar drift paths
for the two data sets and concluded that any changes in the system only
significantly affected the drift speed.

@@

The second method is to compare the experimental ADCs to the energy deposited in 
GEANT4 by similar simulated tracks (using the same elastic events than for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charges along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the data acquisition (DAQ) features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
figure~\ref{fig:EVENT_adc_tdc}. In this step, the gain for each pad is calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains are refined using correcting factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compare it to the average
deposit recorded by the other pads. 
Results of this second method are shown in Figure~\ref{fig:dedx_p_exp_2nd},
from these we concluded that the gains of the 
second method match best the theoretical lines.

@@

\section{Performance Studies}\label{sec_perfor}
The previous discussions have shown that the two modules of the RTPC have 
slightly different yield, however this yield should not be linked to a 
different performance of the RTPC. The differences are mainly due to 
complicated convolution of CLAS and the RTPC acceptances. We measured the 
efficiency of the RTPC using the elastic scattering on $^4$He, where the 
exclusive elastic yield is compared to the inclusive yield for the two modules 
of the detector. Here the inclusive yields are the number of electrons in the 
elastic invariant mass peak whose corresponding elastically scattered $^4$He 
would have been in the acceptance of the RTPC, and the exclusive yields require 
the additional detection of the $^4$He. As a result, we found that the left and 
the right modules have similar efficiencies except near the target windows, as 
shown in figure \ref{fig:rtpc_eff}.

@@

We reported on the construction, operation and calibration of a small RTPC 
designed to measure helium-4 nuclei in high rate environnement. The operation
of the detector was successful and allowed to detect helium nuclei at a rate
of several kHz in coincidence with the detection of high energy electrons and 
photons in the CLAS spectrometer.

@@

@@

\bibitem{BONUS}
   S. Tkachenko et al., Measurement of the nearly free neutron structure 
   function using spectator tagging in inelastic $^{2}H(e,e'p)X$ scattering 
   with CLAS,	Phys. Rev. C 89, 045206 (2014).

@@

\bibitem{BONUS-NIM}
Howard C. Fenker et al., BoNuS: Development and Use of a Radial TPC using Cylindrical GEMs, Nucl. Instrum. Meth. A592, 273-286 (2008)
\end{thebibliography}

>>>>

The CLAS RTPC is a 250~mm long cylinder of 158~mm diameter, leaving just enough 
room to fit pre-amplifiers between the RTPC outer shell and the solenoid. The 
electromagnetic field is directed perpendicularly to the beam direction, 
such that drifting electrons are pushed away from the beam line. These electrons 
are amplified by three layers of curved Gas Electron Multiplier 
(GEM)~\cite{gem_sauli} and detected by the readout system 
on the external shell of the detector as illustrated in the figure~\ref{fig:RTPC_1_4}. The CLAS 
RTPC is segmented into two halves with independent GEM amplification systems that 
cover about 80\% of the azimuthal angle.

We detail here the different regions shown in figure~\ref{fig:RTPC_1_4} starting 
from the beam line and towards larger radius:\\
\begin{itemize}
   \item First, the 6 atm helium-4 target extends along the beam line axis, it 
      is 284~mm long with a 3~mm radius wall made of a 27-$\mu$m-thick Kapton.
   \item The first gas gap covers a radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o ller electrons scattered by the beam. This 
      region is surrounded by a 4 $\mu$m thick window made of aluminized Mylar 
      is electronically grounded.
   \item The second gap region extends between 20~mm and 30~mm and is filled with a 
      gas mixture of 80$\%$ neon and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar, which 
      serves as the cathode and is set at 4260~V.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away 
      from the beam axis. The average electric field in this region is about 550~V/cm 
      and perpendicular to the beam.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied to each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Preamplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

Due to the non perfect experimental conditions, in particular variations in the 
gas mixture~\footnote{It was noted during the experiment that leaks appeared
and the gas flow had to be increased in few occasions to compensate for this.}, 
the drift speed changes during the three months long experimental run.  
Figure \ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for 
individual runs (approximately 2 hours long). We observe a significant 
variation of the drift speed over time and accounted for it in the drift 
speed used for the track reconstruction.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. Software exists to calculate the drift paths, in particular
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, gas composition is far from stable in the 
chamber. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.20 and 1.27 GeV) to enhance the cross section of the elastic 
scatering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematic allows to calculate the Helium nucleus kinematic. It is
by comparing this calculated kinematic to the measured one that we fine tune
the drift paths independently of our knowledge of the exact conditions in the 
chamber.

The drift paths are ajusted using a set of identified elastic events. We generate 
a helium nucleus in a GEANT4 simulation~\cite{GEANT4} of our RTPC for each event
based on the measured electron kinematic. Then, we compare the simulated trajectory of 
the Helium nuclei to the hits measured in the chamber. Because of the 
magnetic field, the drift paths are not linear in the RTPC. So to perform the extraction, 
we make a first approximation with a linear dependence between the radius of 
emission and the time of detection, and then refine our result. As it happens, 
the curvature is minimal and this process converge already on the second 
iteration.

At the end of the extraction procedure, the azimuthal difference between the detection pad and 
the ionization point ($\Delta\phi$) is extracted as a function of time. 
In figure~\ref{fig:DELTA_PHI_TDC}, we show the resulting data points for one 
bin, where the drift paths is easily identified and eventually fitted for 
implementation in our reconstruction codes.

@@

To verify the stability of the drift paths, this procedure was carried out 
using both the 1.204 GeV data from the beginning of the run period and the 
1.269 GeV data from the end of the run period (shown in blue on 
figure~\ref{fig:Drift_run_number_1}). Interestingly, we found very similar drift paths
for the two data sets and concluded that any changes in the system only
significantly affected the drift speed.

@@

The second method is to compare the experimental ADCs to the energy deposited in 
GEANT4 by similar simulated tracks (using the same elastic events than for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charges along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the data acquisition (DAQ) features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Figure~\ref{fig:EVENT_adc_tdc}. In this step, the gain for each pad is calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains are refined using correcting factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compare it to the average
deposit recorded by the other pads. 
Results of this second method are shown in Figure~\ref{fig:dedx_p_exp_2nd},
from these we concluded that the gains of the 
second method match best the theoretical lines.

@@

\section{Performance Studies}\label{sec_perfor}

The previous discussions have shown that the two modules of the RTPC have 
slightly different behavior and give also different particle yield. However, 
this should not necessarily be linked to a different performance of the RTPC. 
Indeed, there is a complicated convolution of CLAS and the RTPC acceptances 
that can significantly affect the yields and the kinematic of the detected 
particles. To clarify this question, we measured the 
efficiency of the RTPC using the elastic scattering on $^4$He, by comparing the 
 the inclusive yield based on electron detection only to the exclusive elastic 
yield where we also detect the Helium recoil. We present in 
Figure~\ref{fig:rtpc_eff} the results for the two modules 
of the detector. We observe that the left and 
the right modules have similar efficiencies except near the target windows, as 
shown in figure \ref{fig:rtpc_eff}.

@@

We reported on the construction, operation and calibration of a small GEM based RTPC 
designed to measure helium-4 nuclei in a high rate environnement. The operation
of the detector was successful and allowed to detect helium nuclei at a rate
of 3.1 kHz in coincidence with the detection of high energy electrons and 
photons in the CLAS spectrometer.

@@

\bibitem{BONUS-NIM}
   Howard C. Fenker et al., BoNuS: Development and Use of a Radial TPC using 
   Cylindrical GEMs, Nucl. Instrum. Meth. A592, 273-286 (2008)

@@

@@

\end{thebibliography}

### 2017-06-02 eg6-letters/NIM/author_list.tex
some fixing
Edits from RD to NIM

\author{R.~Dupr\'e}   \affiliation{Argonne National Laboratory, Argonne IL 60439, USA}\affiliation{Institut de Physique Nucl$\acute{e}$aire, CNRS/IN2P3 and Universit$\acute{e}$ Paris Sud, Orsay, France}
\author{N.~Baltzell}  \affiliation{Argonne National Laboratory, Argonne IL 60439, USA}\affiliation{Jefferson Laboratory, Newport News, VA 230606, USA}
\author{K.~Hafidi}    \affiliation{Argonne National Laboratory, Argonne IL 60439, USA}
\author{M.~Hattawy}   \affiliation{Argonne National Laboratory, Argonne IL 60439, USA}\affiliation{Institut de Physique Nucl$\acute{e}$aire, CNRS/IN2P3 and Universit$\acute{e}$ Paris Sud, Orsay, France}
\author{S.~Stepanyan} \affiliation{Jefferson Laboratory, Newport News, VA 230606, USA}
%\author{} \affiliation{}
%

\collaboration{The JLab-CLAS Collaboration} 
%
\vskip 0.25cm

>>>>

\author{R.~Dupr\'e}   \affiliation{Argonne National Laboratory, Argonne IL 60439, USA}\affiliation{Institut de Physique Nucl$\acute{e}$aire, CNRS/IN2P3 and Universit$\acute{e}$ Paris Sud, Orsay, France}
\author{S.~Stepanyan} \affiliation{Jefferson Laboratory, Newport News, VA 230606, USA}
\author{M.~Hattawy}   \affiliation{Argonne National Laboratory, Argonne IL 60439, USA}\affiliation{Institut de Physique Nucl$\acute{e}$aire, CNRS/IN2P3 and Universit$\acute{e}$ Paris Sud, Orsay, France}
\author{N.~Baltzell}  \affiliation{Argonne National Laboratory, Argonne IL 60439, USA}\affiliation{Jefferson Laboratory, Newport News, VA 230606, USA}
\author{S.~Bueltmann} \affiliation{Old Dominion University, Norfolk, VA 23529, United States}
\author{A.~Celentano} \affiliation{INFN, Sezione di Genova, 16146 Genova, Italy}
\author{R.~De~Vita} \affiliation{INFN, Sezione di Genova, 16146 Genova, Italy}
\author{K.~Hafidi}    \affiliation{Argonne National Laboratory, Argonne IL 60439, USA}
\author{S.~Kuhn} \affiliation{Old Dominion University, Norfolk, VA 23529, United States}

%\author{} \affiliation{}
%

\collaboration{The CLAS Collaboration} 
%
%\vskip 2.0cm
